Columnar VCF using Zarr

### 2024-03-05 paper.tex
Very quick cut of non-relevant text from sgkit paper draft

% Additional packages used for code formatting
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
% Local definitions to systematise tool naming
\newcommand{\sgapi}[1]{\texttt{#1}}
\newcommand{\toolname}[1]{\texttt{#1}}
\newcommand{\sgkit}{\texttt{sgkit}}

@@

\title{Scalable analysis of genetic variation data in Python}

@@

%% NOTE: this is much too long currently, but keeping for now so we 
% can see which stuff migrates to the intro
\begin{abstract}
\textbf{Background:}
Vast quantities of genetic variation data have been collected to 
address fundamental scientific questions across multiple disciplines.
Genetic variation datasets are usually modelled as a ``genotype matrix'',
where rows denote positions on the genome where variation occurs and 
columns represent the state of sequenced samples at that position.
The largest dataset available today, the UK Biobank, 
has a genotype matrix of approximately 1,000,000,000 rows and 500,000 columns,
along with metadata and quality control information of similar dimension.
Processing such matrices on a single computer is not feasible,
and computations must be distributed to complete in a reasonable timeframe.
Classical methods in genomics are inherently single-computer focused, 
oriented around either streaming data through Unix pipelines or 
storing the entire genotype matrix in memory.
The standard file format for genetic variation data (VCF: Variant Call Format)
is also fundamentally unsuited to distributed computation and is a severe 
bottleneck at the Biobank scale.
Researchers either perform ad-hoc distribution by splitting across VCF files
and dispatching jobs using workflow managers, 
or convert data into the undocumented and/or proprietary formats 
used by the few specialised tools that do support automatic distribution
of calculations on a genotype matrix.

\textbf{Results:} We present sgkit, a scalable toolkit for processing genetic 
variation data built on widely-used technologies from the Python data-science
ecosystem. A key part of this approach is to use Zarr, a cloud-native
format for storing multi-dimensional numerical data, widely used in the
across the sciences to store petabyte-scale datasets, and with 
multiple implementations.  We show that data stored in this format has 
comparable compression levels to specialised methods to compress the 
genotype matrix. We show that performing calculations on the genotype
matrix on a single computer is much more efficient at scale using sgkit than standard 
approaches based on VCF, and is comparable with state-of-the-art methods
using compiled languages. Moreover, sgkit can also natively distribute
computation over a cluster using Dask, with no changes required to client code.
We show what sgkit is competitive with Hail (the most popular package for distributed
analysis of genetic variation data) in terms of performance, but is 
much more flexible and extensible. We demonstrate the 
current capabilities of sgkit to perform dataset quality control,
and applications in statistical, population, quantitative and phylo genetics.

\textbf{Conclusions:} 
Sgkit enables the interactive analysis of Biobank-scale genetic variation
datasets via familiar and powerful Python data science tools, boosting
researcher productivity. Leveraging these tools also substantially 
reduces software development effort for creating new methods.
The vcf-zarr specification may additionally provide a useful starting 
point for defining a standard computational format for genetic variation data.
Such a standard, based on the language-agnostic Zarr format, would enable a
diverse ecosystem of tools to efficiently access subsets of the data without
copying, and substantially reduce research costs.
\end{abstract}

@@

\section{Background}
[TODO rewrite this based on the first paragraph of the abstract above 
when cutting it down.]
The study of genetics is currently transforming into a true big-data science.
Since the Human Genome Project, genomics has required working with
signicant volumes of data and the specialised methods
required to transform the raw biological signals into analysable data,
such as assembly[cite], mapping[cite] and variant calling[cite] are compute intensive.
However, beyond these initial steps required to generate the data, the actual
analysis of the genetic variation could typically be performed on a single
computer, and often with ad-hoc software written using shell utilities
and scripting languages such as Perl.
From the Human Genome Project, through 1000 Genomes, and now with
population scale genomics datasets such as UK Biobank, GeL, etc,
the requirements for analysing the data have far outgrown these
methods. Simple queries can require minutes to complete, which
has a negative effect on researcher productivity, as well as limiting
the analyses that can be performed. The financial and 
environmental~\citep{grealey2022carbon,lannelongue2023greener} cost of working
with such datasets is also substantial.

[Summary of what we do and paper roadmap]

In this paper we present sgkit, a Python library for large-scale genetic data
analysis that supports efficient and convenient distributed computation based
on open data formats and established technologies. We begin by examining the
fundamental question of how we can store observed genotypes for millions of
samples and hundreds of millions of variant sites in a way that is compact,
supports efficient retrieval for subsets of both samples and sites, and leads
to efficient distributed computation. We demonstrate that the straightforward
approach of storing genotypes in compressed rectangular chunks using the Zarr
library enables efficient access by sample, and yields compression levels and
computation speeds similar to specialised methods for genetic data. These
chunks also provide
a natural unit % FIXME "unit" is the wrong word
to parallelise work over, and by using the Dask library sgkit can easily and
transparently execute tasks concurrently within a single computer or across a
large cluster. Concurrent and distributed programming is notoriously difficult,
but Dask's high-level array-oriented operations, coupled with just-in-time
(JIT) compilation via numba~\citep{lam2015numba}, makes it straightforward to
write performant distributed code in Python.
Storing and processing genotype data is of course only one part of
genetic data analysis, and real-world analyses require large volumes
of additional data associated with variants, samples and genotypes calls.
To help manage this complexity
the basic unit of sgkit's interface is an Xarray
dataset~\citep{hoyer2017xarray}, which provides a flexible container
for labelled multi-dimensional arrays. Thus, arbitrary numerical data
can be associated with any of the relevant dimensions, and stored
efficiently in chunked Zarr format.

Using these fundamental building blocks sgkit currently provides
functionality across four key themes of general quality-control,
statistical, population and quantitative genetics. In
each case we briefly discuss available software, and showcase
the functionality via case-studies.
% Note: wishlist, just writing down some stuff that sounds good!
% Also four-key themes doesn't quite fit with 5 sections mentioned
% below - it's a WIP.
We first demonstrate interactive quality-control of [large dataset]
on [cluster], demonstrating how users can quickly summarise
large-scale data in a Jupyter notebook. We then illustrate the
statistical genetics functionality by performing a GWAS
on [something]. We showcase support for
population genetic data by [doing something from one of Alistairs
notebooks?].
We then demonstrate support for quantitative genetics
applications by applying relatedness matrix functions to [something].
% Note: it would be quite neat if we could say this is support
% for phylogenetics operations. Coupled with upstream SeqData
% we might actually be able to do this for some public SARS-CoV-2
% data or something, and make a neighbour joining tree using
% scipy (which is a terrible method, so would have to fully
% acknowledge).
Finally, demonstrate sgkit's support for heterogenous computing
and the general potential for using GPUs by computing a
pairwise distance matrix for [something], and demonstrate
X-fold speedup using our GPU implementation.

% Maybe this belongs in the discussion - we can leave it here for now.
Sgkit builds on the four key technologies of Zarr, Dask, Numba and Xarray,
widely used throughout the sciences, to bring efficient large-scale
genetics computation within the reach of many.
Sgkit is by design an open ecosystem.
The data model is well-documented,
based on open standards, and extensible to include arbitrary additional
variables and dimensions. Users are not restricted to a limited palate
of analyses, but have the full power of the Python data-science stack
available, and can write their own operations in terms of distributed
array operations or optimised numerical code. Finally, sgkit's
development model is open, with a codebase that is specifically
designed for ease of contribution,
% This is horrible, put in something better later.
and development processes informed
by combined decades of experience in open-source projects.

@@

\subsection{Overview of narrative for draft}
This subsection is a high-level summary of the main sections and
their narrative connections. \textbf{The section is for authors only and
will be deleted
once the sections themselves have matured reasonably}

@@

\paragraph{Distributed computation}
Why do we need distributed compute? Summary of current
approaches (chunk up along the genome using HPC or workflow
managers). Advantages of true distributed compute, MapReduce
etc. Summary of earlier efforts like ADAM.
Discuss Hail. It's awesome, has enabled current
generation of mega-scale projects. Drawbacks are the
architecture has emphasised performance etc over adaptability
and simplicity. File format is undocumented, etc.
Discuss why the sgkit architecture is suited to distributed
computation. Discuss Zarr's use in cloud stores, and existing
deployments. Discuss Dask.
We do not attempt a head-to-head performance comparison with
Hail here, but defer to Liangde's thesis~\citep{li2022efficient}
with brief summary.
Forward ref to following sections on applications to demonstate
sgkit in a distributed context.

\paragraph{Quality control of large datasets}
Outline the importance of QC, and existing approaches.
Summarise sgkit's general QC capabilities.
We demonstrate sgkit on quality control operations on a large
real dataset. We convert large messy VCF with lots of additional
fields. Quote file size numbers. We do some QC operations on
the dataset, using a good size distributed cluster, using
some of the additional fields. Quote times. Discuss Xarray
here, and possibly illustrate the labelled dimensioned arrays
here via a figure (perhaps a screenshot/pdf conversion of
a Jupyter notebook rendering of the big messy dataset,
showing the various fields).

\paragraph{Statistical genetics}
Quick summary of statgen and existing toolchain.
Summarise sgkit's statgen capabilities.
Demonstrate sgkit doing a GWAS on real data using distributed cluster.
Quote numbers.

\paragraph{Population genetics}
Quick summary of popgen and existing toolchain.
Summarise sgkit's popgen capabilities.
Demonstrate
sgkit doing popgen analysis based on MalariaGEN data (on cluster?)
Quote numbers.

\paragraph{Quantitative genetics}
Quick summary of quantgen and existing toolchain.
Summarise sgkit's quantgen capabilities. Emphasise the importance
of mixed ploidy support.
Demonstrate sgkit doing some relatedness matrix analyses.
Quote numbers, possibly relative to existing tools.

\paragraph{Phylogenetics}
Quick summary of phylogenetics and existing toolchain.
Phylo is not one of sgkit's core capabilities, but there's no
reason it can't support important. Workflows are also becoming
VCF-based, and massive datasets like SC2 would definitely
benefit from the ability to distribute.

We demonstrate the potential by doing computing pairwise
distance matrix from XX samples, and creating a neighbour
joining tree using scipy (which we emphasise is a terrible
idea, but still). We showcase sgkit's support for heterogenous
computing by getting an X-fold speedup on a GPU.

@@

\begin{figure}
    \begin{tabular}{c}
% FIXME temporarily copying this in here until Tim's tikz code arrives.
% https://github.com/pystatgen/sgkit-publication/issues/46
\includegraphics[width=7cm]{diagrams/tmp-illustration.png} \\
\includegraphics[width=7cm]{figures/data-scaling}
\end{tabular}
\caption{Chunked compressed storage of the genotype matrix using Zarr.
(A) Schematic of the chunked array storage used in sgkit
(B) Compression performance compared with BCF, savvy and genozip
based on simulated data (see text for details).
\label{fig-data-storage}}
\end{figure}

@@

\subsection{Quality control of large datasets}
Outline the importance of QC, and existing approaches.
Summarise sgkit's general QC capabilities.
We demonstrate sgkit on quality control operations on a large
real dataset. We convert large messy VCF with lots of additional
fields. Quote file size numbers. We do some QC operations on
the dataset, using a good size distributed cluster, using
some of the additional fields. Quote times. Discuss Xarray
here, and possibly illustrate the labelled dimensioned arrays
here via a figure (perhaps a screenshot/pdf conversion of
a Jupyter notebook rendering of the big messy dataset,
showing the various fields).

\subsection{Statistical Genetics}
Statistical genetics is largely concerned with disease associations
in humans, and the success of genome wide association studies (GWAS) has led
to explosive growth in the field~\citep{uffelmann2021genome}.
Sample size is the single most
important factor in driving statistical power, and
genotype array data has been collected for millions of samples
[more explanation and citations].
GWAS typically use genotype array data because of
its continuing cost-effectiveness, although there has been
exome~\citep{lek2016analysis,backman2021exome}
and large-scale whole genome sequence data is becoming increasingly
available~\citep{halldorsson2022sequences,uk2023whole}.
Genotype array data is usually imputed using software such
as BEAGLE~\citep{browning2018one} or IMPUTE~\citep{howie2011genotype}.
% Mention phasing?

The \toolname{plink}~\citep{purcell2007plink} package is the most
commonly used software for GWAS and has a large range of functionality
required for quality control, association testing and analysis.
More recently software packages have emerged specifically designed
to deal with the very large sample sizes that are becoming
more prevalant, for example SAIGE~\citep{zhou2018efficiently}
Bolt-LMM~\citep{loh2015efficient}
and
REGENIE~\citep{mbatchou2021computationally} are designed to
detect associations in very large cohorts.
The Hail software [citation] has been used [to do lots of things]
[Mention qgg somewhere\citep{rohde2020qgg}]

Because of the relative homogenity of the data sources and
required tasks, programs like \toolname{plink} are able to provide
an end-to-end analysis pipline from the source genotype data, quality
control and associations via Manhattan plots.

\subsection{Population Genetics}
[Things to work in here:
1) Classically popgen is concerned with allele frequencies, and so
used a variety of different data types to access this.
2) Data was at a pretty small scale, so you could do the entire
analysis within a single monolithic package like Arlequin.
3) people have been transitioning to WGS data recently.
]

Population geneticists study evolution at the population scale,
characterising the evolutionary forces shaping the diversity
within and between populations. Compared to statistical genetics,
which is largely concerned with humans, population geneticists
study a broad range of organisms across the tree of life, both well
characterised ``model'' species, and those for which the most basic
genetic information (such as the sizes and number of chromosomes,
and ploidy level) must be determined.
Sample sizes are also much smaller,
with dozens to a few hundred genomes per species being typical.
However, projects such as Darwin Tree of
Life~\citep{darwin2022sequence} are currently sequencing tens of
thousands of different species.

% JK: I'm making this up - need to get something concrete to back it up.
% Is there any review paper summarising computation in popgen?
[THIS IS A ROUGH DRAFT: I'm just getting in a paragraph here with citations
to tools/packages that are important in PopGen. We can get a narrative
through it later.]
See \citep{excoffier2006computer} for a review of computational methods
in classical population genetics.
Because of the relatively small volumes of data and the variety
of different organisms, statistical analysis of population genetic
data has tended to be done by custom scripts. Tools such
as \toolname{plink}~\citep{purcell2007plink}
are often not applicable because of the built-in
assumptions that are appropriate only for humans and other model organisms.
Programs such as Arlequin~\citep{excoffier2005arlequin,excoffier2010arlequin}
contain a large suite of features, but can only be accessed by GUI
or command line interfaces.
BioPerl~\citep{stajich2002bioperl} and BioPython~\citep{cock2009biopython}
are more concerned with alignments and providing access to command-line
programs, and are oriented towards flexibility rather than
performance.
The \toolname{scikit-allel}~\citep{miles2023scikit}
and \toolname{pylibseq}~\citep{thornton2003libsequence}
libraries provide efficient programmatic
access to a rich suite of population genetic statistics, but have limited
support for parallelisation.

STRUCTURE~\citep{pritchard2000inference}
ADMIXTOOLS~\citep{patterson2012ancient}

% Discussion of simulated data and tree-based stuff.
Population geneticists are often interested in comparing with simulated data.
Until recently, this was done by processing
custom text formats output by programs such as
\toolname{ms}~\citep{hudson2002generating} and accompanying
scripts to compute basic statistics.
% JK: This is a bit of a self-cite fest here, but I think people would
% wonder what the connection between sgkit and tskit is if we don't
% explain it. Happy to discuss if anyone doesn't like it, though.
The \toolname{msprime}~\citep{kelleher2016efficient} simulator took a
different approach by providing a Python API which gives
access to the underlying genealogical trees, which often allow
statistics to be calculated more efficiently than standard
matrix-based approaches~\citep{ralph2020efficiently}.
These methods have been subsequently generalised to
other forms of
simulation~\citep{kelleher2018efficient,haller2018tree}
as well as being inferred from observed variation
data~\citep{kelleher2019inferring,wohns2022unified},
supported by the open-source \toolname{tskit}
library~\citep{ralph2020efficiently}.
See~\cite{baumdicker2021efficient} for further discussion on
computation in population genetic simulations, and the
\texttt{tskit} software ecosystem.
% Need to say this somewhere, or people will wonder. This is a
% rough draft
While tree-based methods are excellent for simulated data
and attractive for real-data analysis, they are not a solution
to all problems. Firstly, not all datasets are appropriate
for tree inference and secondly, before we infer trees we must
first perform extensive quality control of the raw data, necessarily
in variant-matrix form.

Sgkit provides a number of methods for computing statistics in population
genetics. All methods work on an Xarray dataset that has a number of well-known
variables defined [TODO: see section...], which for population genetics are the
genotype calls, stored in the \sgapi{call\_genotype} variable. The methods return a
new dataset containing variables that store the computed statistics.

Before running the methods, the dataset is usually divided into windows along
the genome, using the \sgapi{window\_by} functions, which tell sgkit to produce
per-window statistics. For example, \sgapi{window\_by\_position} creates windows that
are a fixed number of base pairs, while \sgapi{window\_by\_interval} creates windows
corresponding to arbitrary user-defined intervals.

It's common in population genetics to group samples into populations, which in
sgkit are referred to as \emph{cohorts}. There are two types of statistics: one-way
statistics where there is a single statistic for each cohort, and multi-way
statistics where there is a statistic between each pair, triple, etc of
cohorts. [TODO: do we need to say how cohorts are defined?]

The methods for one-way statistics include \sgapi{diversity} for computing mean
genetic diversity, \sgapi{Tajimas\_D} for computing Tajima’s D, and
\sgapi{Garud\_H} for
computing the H1, H12, H123 and H2/H1 statistics~\citep{garud2015recent}.

The methods for multi-way statistics include \sgapi{divergence} and
\sgapi{Fst} for
computing mean genetic divergence and $F_{ST}$ (respectively) between pairs of
cohorts, and \sgapi{pbs} for computing the population branching statistic between
cohort triples.

We converted phased Ag1000G hypotype data in Zarr format
%[@https://www.malariagen.net/data/ag1000g-phase-2-ar1]
to sgkit's Zarr format
using the \sgapi{read\_scikit\_allel\_vcfzarr} function.
The data contained 1,164
samples at 39,604,636 sites, and was [TODO] MB on disk before conversion, and
[TODO] MB after conversion to sgkit's Zarr format. Data for the X chromosome
was discarded since it was not available for all samples. The conversion took
[TODO: X minutes Y seconds], including a postprocessing ``rechunk'' step to
ensure that the data was suitably chunked for the subsequent analysis.

We grouped samples into 15 cohorts based on location using simple Pandas and
Xarray data manipulation functions. We then set up half-overlapping windows
along the genome using sgkit's \sgapi{window\_by\_position} function with a fixed
window size of 6000 base pairs, and a step size of 3000 base pairs.

We computed H statistics for one cohort using the \sgapi{Garud\_H} function in sgkit.
This operation took [TODO: X minutes Y seconds]. Finally, we verified that the
output was concordant with scikit-allel. [TODO: how hard to reproduce the
scikit-allel visualization?]

[TODO: summarize the biologically interesting thing here]

\subsection{Quantitative Genetics}
[TODO lacking a lot of detail here in terms of compute and software.
https://github.com/pystatgen/sgkit-publication/issues/26]

Quantitative genetics is classically concerned with understanding
the link between continuously variable quantitative traits
with complex genetics~\citep{hill2010understanding}.
Through decades of selective breeding, quantitative genetics methods have
led to X-fold increases in milk yield [etc, citations].
They are also applied to
natural populations~\citep{wilson2010ecologist}
Quantitative genetics methods are used across a wide variety of
organisms, with many species of argicultural interest having
variable ploidy levels etc.

Much of classical quantitative genetics uses observed pedigree structure
in mixed models to partition the variance of phenotypes, without
reference to any genomic data.
Fitting generalised linear mixed models (LMMs) using
software such as ASReml~\citep{butler2009asreml}
and MCMCglmm~\citep{hadfield2010mcmc} require only the pedigree
information and trait data.
Genetic marker data~\citep{bernardo2008molecular} of various
types (RFLPs, microsats etc) required statistical models.
BLUPs include genetic information~\citep{endelman2011ridge}.

Many packages are available in R [cites]. Because much of the computation
in quantitative genetics has been around solving
general statistical models using linear algebra, R is very good.
However, with larger and larger samples sizes, particularly
in agricultural applications.
For example,
the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, embedded in multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{cesarani2022mulibreed}.
[Methods are hitting a computational wall]

Much of the functionality required for quantitative genetics work
is therefore provided by access to efficient and scalable linear algebra
routines (a core element of the PyData ecosystem), the ability
to access genetic variation data from sources such as plink and VCF,
and some specialised routines to handle pedigree data and to
estimate relatedness matrices. Sgkit provides [this], with a particular
focus on supporting autopolyploids.

@@

\item TODO several missing paragraphs here discussing the
benefits of the other layers of the stack.

\item Closing. Sgkit, built on the fundamental idea of storing
data in a simple, efficient and distributable fashion, would
hopefully be one element of this compute ecosystem.
Xarray, Dask and numba provide powerful tools to make users
more productive, and enable the interactive analysis of large
scale variation data. We have illustated the current functionality
through examples here across X-genetics, but there is much
potential for more. The library is specifically designed to make
contribution easy, and we welcome people to the open-source
community.

@@

%%% OLD TEXT
% \begin{itemize}

% \item While API based approaches such as htsget~\citep{kelleher2019htsget}
% and more generally federated access to data~\citep{rehm2021ga4gh}
% are an important part of the future of genomics, we must still pay
% attention to how efficient \emph{local} compute is. We need to
% store and exchange data in an efficient and easily accessible
% format. [Could talk about
% how early GA4GH APIs failed because they didn't really consider
% efficiency? Converting VCF to JSON was never going to be efficient?
% Might not be worth getting into]

% \item Users of sgkit will be able to perform data manipulation, analysis, and
% visualization using familiar PyData tools, with only domain-specific methods
% requiring domain-specific APIs. They will also benefit from interactive
% performance for medium-sized data and the ability to scale out to biobank-scale
% data within the same library.

% \item
% Methods developers won’t have to reimplement undifferentiated functionality
% like defining a custom data model, reading from files into that data model, and
% validating data. Instead, they can implement their novel methods against
% sgkit’s data model and instantly gain adjacent functionality and accessibility
% to sgkit’s user community. This is well illustrated by our implementation of
% the state-of-the-art REGENIE algorithm, which requires less than 1000 lines of
% Python code (including documentation) and has comparable performance to the
% original C++. Further, many key methods in statistical genetics operate on GWAS
% summary statistics, and integrating individual level data with external summary
% statistics is an increasingly important strategy which other toolkits have not
% yet formalized.

% \item sgkit makes it easier for researchers to
% make their whole analysis pipeline available in a reproducible Jupyter
% notebook. Most statistical genetics papers describe the collection of
% command-line utilities used in their methods section, and a select few make a
% workflow connecting the command-line utilities available. Users of sgkit can
% work within a Jupyter notebook without having to call out to command-line
% utilities, making the journey from raw data to scientific findings easier to
% represent in a single notebook.

% \end{itemize}

@@

\subsection{Scaling up interactive analysis}
The storage and analysis of datasets consisting of hundreds of
thousands of genomes presents major challenges.
Fig.~\ref{fig-data-storage} shows the performance of \toolname{sgkit}
in terms of storage space and computation time  on a simple benchmark (see
below for details) with increasing
numbers of samples, contrasting with the standard BCF file format
and \toolname{bcftools} utility.
The dataset used is a subset of the
ultra-realistic simulations of 1.4 million French-Canadians
provided by~\citet{anderson2023genes}, which is freely avilable
and has been shown to
accurately reflect fine-scaled demographic structure.
Although
simulations can never capture the full richness of real data,
this resource certainly reflects the patterns of allele frequency
and correlation structure that we would expect in large-scale
human studies, and provides a useful and reproducable benchmark.
Note that as we increase the sample size, the number of variant
sites also increases [following basic popgen principles?].

The scale of present-day genomic data means that compact
storage is of paramount importance, and reductions in storage space
can make a significant difference in overall costs.
Fig.~\ref{fig-data-storage}A shows that the
compressed-chunked-array Zarr format used by \toolname{sgkit} results
in significantly smaller file-sizes than BCF (encoded using
\texttt{bcftools view -O b}; we used \toolname{bcftools}
version 1.18 throughout). At the extreme, when we have 1M
samples, we require XGB for the BCF file and a total of YGB
for the \toolname{sgkit} version. Note that under the default
directory based Zarr store this results in XX files of around
y KB each to store the chunks. Different Zarr storage engines
can be used, varying from a ZipFile which keeps the entire
dataset in a single file to a cloud store [details] which stores
chunks in [a cloud store]. The chunk size is a parameter and
can be tuned. [bit more detail here].

The scale of modern genomic datasets also means that calculations must
be parallelised. This is illustrated in Fig.~\ref{fig-whole-matrix-compute}B,
where we compare the performance of a simple calculation requiring
the examination of all genotypes using \toolname{bcftools}
and \texttt{sgkit}. Interactive analysis is only possible through
parallelisation, as optimised single-threaded code
is simply not fast enough.
Although it is relatively straightforward to split calculations across
the genome using tools like \toolname{bcftools}, the process
is quite manual, and often involves working with HPC schedulers
or workflow managers [this is very rough, just seeing where it goes].
Splitting the workload \emph{by sample} however, is much more
difficult. Parallelisation by variant and sample is an inherent
part of \texttt{sgkit}'s design---the variant matrix is chunked
in two dimensions, and Dask transparently maps computations to these
chunks.
Fig.~\ref{fig-whole-matrix-compute}B shows how \texttt{sgkit} scales
on a single server, and how computation time decreases as we use
more threads. In contrast, using threads with
\toolname{bcftools} (via the XXX option) makes almost no difference
because computational model is inherently sequential, and additional threads
are only used to facilitate decompression.
As we see in section XXX,
a major advantage of \texttt{sgkit} is that it not only supports
scaling \emph{up} within a server, but also scaling \emph{out}
across multiple servers.

The ability to
run pre-defined analyses interactively on very large datasets by
transparently distributing across a cluster is a major boost to
researcher productivity. The ability to define one's \emph{own}
analyses using familiar Python data-science tools and libraries,
and to similarly distribute across a cluster (with single-threaded
speed close to native compiled code) is a game-changer [say better].
Fig.~\ref{fig-whole-matrix-compute}C shows the code used in the benchmark
for \toolname{sgkit} in Fig.~\ref{fig-whole-matrix-compute}B. The heavy
computation here is done in the \texttt{variant\_stats(ds)} function,
which computes a set of variables summarising each of the variants
by examining the entire variant matrix.
These variables are mostly one-dimensional, and can thefore
be processed `in-core' using numpy methods very effectively.
For example, we \texttt{variant\_n\_het} the number of heterozygous samples
in Fig.~\ref{fig-whole-matrix-compute}C. For $n=10^6$ we have X million
variants, and the overall size of the data is Y MiB which
comfortably fits in memory. Thus, the
\texttt{get\_prob\_dist} function shown here reproduces the functionality
of the \toolname{bcftools +afdist} plugin, which must be compiled and
requires detailed knowledge of \toolname{htslib} APIs, using standard
Python data science tools, and with similar performance.

Reducing the variant matrix to summaries that are separately stored
is the standard practise when using VCF/BCF files, for precisely the
reason of avoiding the need to examine the entire genotype matrix.
These are most often stored as ``INFO`` fields.
Indeed, \toolname{bcftools +afdist} requires that the \texttt{AC}
INFO field is present (but does not raise an error
if it doesn't). Because of the row-based storage approach of
VCF/BCF, adding any new fields to a file means reading in the entire
file, and writing out a full copy. For our $n=10^6$ samples example,
computing \texttt{AC} value required XX hours.
In contrast, the columnar storage approach (see section XXX) used
in \toolname{sgkit} means that we can \emph{update} an existing
dataset with this derived information, without altering the existing
on-disk information.
Computing the
\texttt{variant\_allele\_counts} variable using
40 threads  required XX minutes, and updating the existing dataset required
0.X seconds.

The shortcomings of VCF as a storage format and computational
substrate are well known, and as we discuss in Section XXX,
storing genetic variation data is a complex issue
and the subject of substantial ongoing research. The point of this
section is not to establish that the approach taken by \toolname{sgkit}
is state-of-the-art, but to contrast the approach we have taken
with VCF (as the standard storage format) and to demonstrate
the scalability and efficiency of \toolname{sgkit} as an analysis
platform.

\subsection{Sgkit design principles}

Openness is a core organising principle of sgkit, both in terms of
development processes and code licensing, and also with respect
to usability. By following the general philosophy of array-oriented
computing, the functionality in sgkit is organised around the
key concept of a ``dataset`` which encapsulates a set of
related ``variables''. The implementation aims to be as agnostic
as possible as to the underlying implementation of these arrays,
which can be stored in-memory as NumPy arrays, in chunked
compressed on-disk storage as Zarr arrays, or ...

% Perhaps this should be mentioned in the QuantGen section?
The data layout used here is compatible with the ChromaX
breeding program simulator~\citep{younis2023chromax}, which
is leveraging other aspects of the PyData ecosystem to
fully utilise modern parallel hardware.

[In later sections we discuss enabling technologies in detail,
here we talk about how we use them from a high-level, and layout
the overall design philosophy. This isn't documentation, rather
it's the forum where we lay out explicitly early design choices
and their rationale.]

VCF conversion is handled using
\toolname{cyvcf2}~\citep{pedersen2017cyvcf2}.
[TODO any citations for these?]
BGEN using \toolname{cbgen}.
Plink support using \toolname{bed\_reader}

[Text from proposal]

The rapid evolution of supporting projects in the PyData ecosystem, including Dask for
distributed processing, Xarray for coordinated operations on groups of labeled
n-dimensional arrays, Zarr for n-dimensional array storage on cloud object
stores, and Numba for hardware acceleration. We’ve designed sgkit from the
ground up to tightly integrate with these libraries so that we can benefit from
the many improvements being made upstream, making the PyData ecosystem
accessible to statistical geneticists.

We've also defined a Zarr representation for efficient storage of this data.
We’ve designed an Xarray dataset-based data model which can be populated from
any of the file formats above, and against which library methods are
implemented. These methods include variant and sample quality control,
population structure analysis, genome-wide selection scans, and genome-wide
association analyses, including a novel implementation of the recently
developed and state-of-the-art REGENIE algorithm.

\section{Enabling technologies}

[High-level summary of where things are in terms of data analysis
in genomics, pointing forwards to
the individual use-cases of popgen, statgen etc to discuss the
individual tools.]

Other sciences have been working at the petabyte scale for some time.
[I don't really know the background here, it'll require some research.
Astronomics, geosciences etc have all been working with such large
datasets, and been using Python to do so, successfullly. Concrete
examples]

While some aspects of genomics data are unique and require specialised
tooling, by the time that variants have been called and the we wish
to work on the final stages of analysis, we are usually
working with large n-dimensional arrays of numerical data. Such
datasets are precisely what a large and active community of
developers have been working on in the PyData Ecosystem. The goal
of sgkit is to bring these tools to genomics, using a
shared platform for analysing large arrays of numerical data. In the
following subsections we discuss the key technologies, and how
they apply to genomics.

@@

\subsubsection{Just-in-time compilation}

Many of the analyses that we wish to perform in genomics are unique,
and the ability to quickly put together complex queries using
high-level tools like shell pipelines and scripting languages
is often a major bonus. However, such approaches are often not
viable at the petabyte scale, where efficient compiled code that
uses available CPU resources effectively are required. However,
writing code in compiled languages such as C, C++ or Java
is far more labour intensive, as well as requiring more specialised
software development experience.

Similarly, [scripting languages such as python are great for
contributing to open source libraries, because there's a much
lower barrier to contribution.]

We use numba because [it's awesome].

\subsubsection{Heterogenous computing}

GPUs are cool, we support them.

@@

\subsection{Reimplementing complex methods}

To showcase the advantages of \sgkit's pure-Python development model, we
discuss some things we've implemented.

\subsection{Functionality}

[Just listing out all the things that have been implemented to try and get
some kind of grip on what we're doing]

\begin{itemize}

\item \sgapi{pc\_relate} estimate genetic relatedness, returns sample x sample
kinship coefficient matrix. \citep{conomos2016model} (genetic Relatedness matrix)

\item \sgapi{Weir\_Goudet\_beta} sample x sample Kinship matrix
\citep{weir2017unified} (genetic Relatedness matrix)

\item \sgapi{genomic\_relationship} Genetic relationship matrix estimated
by either the Van Raden algorithm generalised to
autopolyploids~\citep{vanraden2008efficient,ashraf2016estimating,bilton2020developing}
and the Endelman-Jannink estimator \citep{endelman2011ridge}
(genetic Relatedness matrix)

\item \sgapi{hybrid\_relationship} Hybrid relationship matrix
(AKA the HRM or H-matrix) combining pedigree and genomic information
~\citep{martini2018effect}
Also \sgapi{hybrid\_inverse\_relationship} (Relatedness matrices)

\item \sgapi{pedigree\_inverse\_kinship} Calculate the inverse of the kinship
matrix from pedigree structure.

\item \sgapi{pedigree\_kinship} Calculate the kinship
matrix from pedigree structure.

\item \sgapi{invert\_relationship\_matrix} Compute the inverse of a
relationship matrix (?)

\item \sgapi{pedigree\_contribution} expected genomic contribution of each
sample to each other based on pedigree (Hamilton-Kerr?)

\item \sgapi{pedigree\_inbreeding} Estimate expected per-sample
inbreeding coefficients from pedigree structure.
\citep{hamilton2018computation}

\item \sgapi{identity\_by\_state} Pairwise sample x sample matrix
of IBS probabilities

>>>>

% Additional packages used for code formatting
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}

% Additional packages for data model figure
\usepackage{forest}
\usepackage{tikz}
\usetikzlibrary{quotes,arrows.meta,3d}

% Local definitions to systematise tool naming
\newcommand{\sgapi}[1]{\texttt{#1}}
\newcommand{\toolname}[1]{\texttt{#1}}
\newcommand{\sgkit}{\texttt{sgkit}}

@@

% Macros etc for data model figure
\input{diagrams/data-model-preamble.tex}

% tentative
\title{Columnar VCF using Zarr}

@@

%% NOTE: this is much too long currently, but keeping for now so we
% can see which stuff migrates to the intro
\begin{abstract}
\textbf{Background:}
The largest dataset available today, the UK Biobank,
has a genotype matrix of approximately 1,000,000,000 rows and 500,000 columns,
along with metadata and quality control information of similar dimension.
Processing such matrices on a single computer is not feasible,
and computations must be distributed to complete in a reasonable timeframe.
Classical methods in genomics are inherently single-computer focused,
oriented around either streaming data through Unix pipelines or
storing the entire genotype matrix in memory.
The standard file format for genetic variation data (VCF: Variant Call Format)
is also fundamentally unsuited to distributed computation and is a severe
bottleneck at the Biobank scale.

\textbf{Results:}
key part of this approach is to use Zarr, a cloud-native
format for storing multi-dimensional numerical data, widely used in the
across the sciences to store petabyte-scale datasets, and with
multiple implementations.  We show that data stored in this format has
comparable compression levels to specialised methods to compress the
genotype matrix. We show that performing calculations on the genotype
matrix on a single computer is much more efficient at scale using sgkit than standard
approaches based on VCF, and is comparable with state-of-the-art methods
using compiled languages.

\textbf{Conclusions:}
Such a standard, based on the language-agnostic Zarr format, would enable a
diverse ecosystem of tools to efficiently access subsets of the data without
copying, and substantially reduce research costs.
\end{abstract}

@@

\section{Background}

@@

\subsection{Overview of narrative for draft}

@@

@@

\begin{figure}
    \begin{tabular}{c}
\resizebox{7cm}{!}{\input{diagrams/data-model.tex}} \\
\includegraphics[width=7cm]{figures/data-scaling}
\end{tabular}
\caption{Chunked compressed storage of the genotype matrix using Zarr.
(A) Schematic of the chunked array storage used in sgkit
(B) Compression performance compared with BCF, savvy and genozip
based on simulated data (see text for details).
\label{fig-data-storage}}
\end{figure}

@@

@@

@@

@@

@@

@@
